Making Sequential Programs Deterministic

### 2019-04-06 design.tex

\<OrderNonDet> may only be written on collections and maps.
A map is a dictionary or associative array, such as a hash table.
\OurTypeSystem largely treats a map as a collection of key--value pairs.
Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
the element types can be specified independently of the collection type,
including their qualifiers.

@@

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, etc.
(Rules \rulename{get}, \rulename{iteration}, and \rulename{next element} in \cref{fig-ordernondet-rules}.)
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

@@

\noindent
and no type introduced so far can express this polymorphism.
(The typing is actually even more complex, because if \emph{either}
argument to \<get> is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.  The above examples only show the case where the index
is deterministic.)
The typical type for the list access operation only applies if both
arguments are \<Det>,
and our typing must have that as a special case, which is the first line above.

\begin{figure}
\todo{Give the types for get and set, which are more basic list/array operations,
  at the beginning before iteration.}

@@

    \rulename{iteration}: 
    $
    \forallt{\tau} \<OrderNonDet> \ C \angles{\tau} \rightarrow 
    \<OrderNonDet> \ \<Iterator> \angles{\tau}
    $

@@

    \rulename{next element}:
    $
    \forallt{\kappa\ \beta} \<OrderNonDet>\ \<Iterator>\angles{\kappa \ \beta} \rightarrow
    \<NonDet>\ \beta
    $

@@

    \rulename{hasnext element}:
    $
    \forallt{\tau} \<OrderNonDet>\ \<Iterator>\angles{\tau} \rightarrow
    \<Det> \ \<boolean>
    $

@@

\caption{\<OrderNonDet> Collection rules.  $C$ is a collection class.
    These are partial types indicating the methods' behavior on
    \<OrderNonDet> arguments.  Full types are given in \cref{polymorphism}.}
\todo{These should be method names, not rule names, right?  The figure is
  just giving the types of methods.}
\label{fig-ordernondet-rules}
\end{figure}

@@

$\<length>\ : \<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$

$\<length>\ :\<Det>\ \<String> \rightarrow \<Det> \<Int>$

@@

\begin{tabular}{lll}
  Expression & instantiation & type \\
\<plus(ni, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
\<plus(di, di)> & $\<Det>\ \<String> \rightarrow \<Det> \<Int>$ & \<Det> \\
\<plus(ni, di)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
\<plus(di, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
\end{tabular}

>>>>

\<OrderNonDet> may only be written on collections and maps.
A map is a dictionary or an associative array, such as a hash table.
\OurTypeSystem largely treats a map as a collection of key--value pairs.
Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
the element types can be specified independently of the collection type,
including their qualifiers.

@@

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, etc.
(As shown by types of methods \rulename{get}, \rulename{iterator}, \rulename{next} in \cref{fig-ordernondet-rules})
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

@@

\noindent
and no type introduced so far can express this polymorphism.
(The typing is actually even more complex, because if \emph{either}
argument to \<get> is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.  The above examples only show the case where the index
is deterministic.)
The typical type for the list access operation only applies if both
arguments are \<Det>,
and our typing must have that as a special case, which is the last line above.

\begin{figure}

@@

    \rulename{iterator}: 
    $
    \forallt{\tau} \<OrderNonDet> \ C \angles{\tau} \rightarrow 
    \<OrderNonDet> \ \<Iterator> \angles{\tau}
    $

@@

    \rulename{next}:
    $
    \forallt{\kappa\ \beta} \<OrderNonDet>\ \<Iterator>\angles{\kappa \ \beta} \rightarrow
    \<NonDet>\ \beta
    $

@@

    \rulename{hasnext}:
    $
    \forallt{\tau} \<OrderNonDet>\ \<Iterator>\angles{\tau} \rightarrow
    \<Det> \ \<boolean>
    $

@@

\caption{\<OrderNonDet> Collection rules.  $C$ is a collection class.
    These are partial types indicating the methods' behavior on
    \<OrderNonDet> arguments.}
%\todo{These should be method names, not rule names, right?  The figure is
%  just giving the types of methods.}
\label{fig-ordernondet-rules}
\end{figure}

@@

$\<length>\ : \<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$

$\<length>\ :\<Det>\ \<String> \rightarrow \<Det> \  \<Int>$

@@

\begin{tabular}{lll}
  Expression & instantiation & type \\
\<plus(ni, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$ & \<NonDet> \\
\<plus(di, di)> & $\<Det>\ \<String> \rightarrow \<Det> \ \<Int>$ & \<Det> \\
\<plus(ni, di)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$ & \<NonDet> \\
\<plus(di, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$ & \<NonDet> \\
\end{tabular}
